Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-10-29 latex/paper.tex
More content: and bibliography

\author{\IEEEauthorblockN{Georgios Gousios}
\IEEEauthorblockA{
Delft University of Technology\\
Delft, The Netherlands\\
Email: G.Gousios@tudelft.nl} 
\and
\IEEEauthorblockN{Erik Meijer}
\IEEEauthorblockA{Microsoft \\
Rendmond, WA, USA\\
Email: H.J.M.Meijer@tudelft.nl} 
}   
\maketitle

@@

  In the last few years, the popularity of functional programming as a way
  of solving computational problems has increased significantly. While most
  computer science curricula do include a course on functional programming,
  in many cases it is disconnected from practical applications, which is
  precisely where functional programming shines. To fill in this gap, we 
  designed a functional programming course that demanded from students to learn
  by experience with real world applications. In this work, we present our
  experiences with conducting this course.
\end{abstract}

@@

@@

  \item To introduce students to basic functional programming concepts, such
    as higher order functions, monads and type inference.

  \item To introduce students to solving data processing problems in a 
    functional way.

@@

The course consisted of a series of lectures

\subsection{Student profile}

The course was elective at the MSc curriculum. Students from all departments in
the Electrical Engineering, Mathematics and Computer science faculty were
allowed to participate. Most students had received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce in the web information systems course). Two of the
students were majoring in computer engineering, which meant that their
programming experience was somewhat limited. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

For organizational reasons, the participation to the course was limited to 15
students.

@@

@@

  \item Multi-source real time data processing. Students worked on a 
    programming language popularity index

  \item Implementation of Haskell constructs in Javascript. Students implemented
    the Haskell prelude (basic functions for list manipulation).

@@

  \item Machine learning algorithms. Students implemented Na\"ive Bayes 
    classifiers and K-means clusters.

@@

@@

@@

We would like to thank\ldots

>>>>

\author{\IEEEauthorblockN{Erik Meijer}
\and
\IEEEauthorblockN{Georgios Gousios}
\IEEEauthorblockA{
Delft University of Technology\\
Delft, The Netherlands\\
Email: {H.J.M.Meijer, G.Gousios, Arie.vanDeursen@tudelft.nl} 
}  
\and
\IEEEauthorblockN{Arie van Deursen}
}
\maketitle

@@

  In the last few years, the popularity of functional programming as a way of
  solving computational problems has increased significantly. While most
  computer science curricula do include a course on functional programming, in
  many cases it is disconnected from practical applications, which is
  precisely where functional programming shines. To fill in this gap, we
  designed a functional programming course that demanded from students to
  learn by experience with real world applications. In this work, we present
  our experiences with conducting this course.

\end{abstract}

@@

\section{Challenges}

While planning the course, we dealt with the following challenges related
to the course's organization and content.

\subsection{Heterogeneous Background of participants}

The functional programming course is elective at the MSc curriculum. Students
from all departments in the Electrical Engineering, Mathematics and Computer
science faculty are allowed to participate. For practical reasons, the
enrollment to the course was limited to 15 participants. This permitted
us to get to know the students individually and to provide them with
personalized support.

From the students that enrolled, most have received formal introduction to
imperative and object oriented programming in their bachelors curriculum, while
through participation to other courses they had limited exposure to functional
programming (i.e. Map/Reduce~\cite{Dean04} in the web information systems
course). Two of the students were majoring in computer engineering, which meant
that their programming experience was restricted. Several students also had work
experience as programmers as part of their industrial placement or through
participation to start up ventures.

The expected diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. While the

\subsection{Choosing the appropriate topics}

Functional programming is arguably the oldest programming paradigm. In it is
purest form, it is based on a minimalistic theoretical background
($\lambda$-calculus~\cite{Baren84}). Relatively recent
formulations~\cite{Meije91, Wadle93} also introduced concepts from category
theory. While the every day use of functional programming does not necessarily
require the programmer to be aware of the theoretical background, understanding
it usually leads to more elegant algorithmic solutions. Functional
programming teaching is also associated with advanced type systems; the flagship
functional programming languages (Haskell and the {\sc ml} family) both offer
very sophisticated support for types. Consequently, teaching the full spectrum
of functional programming theory and techniques would have been impossible in a
half-semester course; instead we decided to focus on practical aspects of data
processing, transformations and state representation using functional techniques.

\subsection{Choosing the appropriate programming language}

@@

  \item To introduce students to basic functional programming concepts, such
    as higher order functions, monads and advanced type systems.

  \item To introduce students to the approach of viewing data processing
    problems as a series of function applications.

@@

The course consisted of a series of lectures, of which two where invited
lectures by an external instructor, laboratories, a series of intermediate
assignments and a final assignment. The students would be evaluated by their
performance on the final assignment only.

@@

\cite{Hutto07}

@@

  \item Multi-source real time data processing. Students worked on a 
    programming language popularity index, based on data from the Github
    event stream and the StackOverflow tag cloud {\sc api}.

  \item Implementation of Haskell constructs in Javascript. Students implemented
    the Haskell Prelude (basic functions for list manipulation).

@@

  \item Implementation of simple constraint solvers.

  \item Machine learning algorithms. Students implemented Na\"ive Bayes 
    classifiers and K-means clustering algorithms.

@@

The assignments were open ended, as there was no fixed set of targets for
each one of them. The deliverables for the assignments were a repository
with the source code and short report describing the solution that the
students came up with. The students were also required to have a working
demo by the end of the course time.

\subsection{The labs}

@@

Teaching the concepts of solving problems algorithmically can be a daunting
task~\cite{Futsc06}; especially more so, when the subjects are already in a
specific mindset. The assumption of the presented course was that imperative and
functional programming are really the two sides of the same coin; by focusing on
how students can apply rather than just be taught concepts of functional
programming using the, typically imperative, languages they already know, allows
them to better appreciate the strong points and weaknesses of each paradigm. In
our experience, the approach has been successful. Both throughout the initial
homework assignments and in their projects, the students demonstrated a high
degree of appreciation of functional programming concepts. Above all, we
believe that it was the hands-on approach through tough homework exercises
employed in this course that allowed students to sharpen their skills and
understand the concepts that were taught during the course.

@@

We would like to thank the course participants for their enthusiasm and
dedication.
